---
title: "Blood Gas Data from NEAq Research"
output: slidy_presentation
---
```{r, include=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)

# load clean data
sliders <- read.csv('~/Dropbox/Blood_Gas_Project/Data/tidy_turtle_data.csv')
complete_sliders <- read.csv('~/Dropbox/Blood_Gas_Project/Data/turtles/focal_turtle_data_zeros.csv')
rays <- read.csv('~/Dropbox/Blood_Gas_Project/Data/tidy_ray_data.csv')
complete_rays <- read.csv('~/Dropbox/Blood_Gas_Project/Data/focal_ray_data.csv')

complete_sliders <- complete_sliders[1:59,2:7]
complete_rays <- complete_rays[1:101,2:8]

# variables gathered to plot all on one graph
sliders_gath <- complete_sliders %>%
  gather(key = "Variable", "Value", 3:6)

rays_gath <- complete_rays %>%
  gather(key = "Variable", "Value", 3:6)

rays_gath <- rays_gath %>%
  filter(!is.na(Value))


sliders_gath <- complete_sliders %>%
  gather(key = "Variable", "Value", 3:6)

write.csv(sliders_gath, file = "sliders_gathered.csv")

```

## 1) What is the dataset? {.smaller}
- Blood samples were collected in 2013 from 10 Turtles (Red-eared sliders), and 10 stingrays (Cownose rays)
- Turtles were selected from Rainforest Reptile Shows in Beverly, MA
- Rays were opportunistically selected from the shark and ray touch tank at the NEAq during routine spine clipping
- blood was analyzed for pH, pCO2, pO2, and lactate values using an iStat analyzer
- each sample was analyzed immediately (T0) and at 5, 10, 15, 45, and 90 minutes following 

```{r, collapse=TRUE}
head(complete_sliders)
head(complete_rays)
```


## 2) What is your question?
- Are there changes in blood acid-base parameters over time?
- Do the blood acid-base parameters different between different syringe types (glass vs plastic)?  
  
*These values are typically measured to assess stress levels of fish, samples are taken on a boat and usually can't be analyzed until they are brought back to a lab.*
```{r}


```

## 3) One visualization that addresses the question
```{r, echo=FALSE}
ggplot(data = sliders_gath, aes(x=Time, y=Value, color=factor(Turtle), by=Turtle)) +
  geom_line() +
  geom_point() +
  facet_wrap(~Variable, scales = "free")
```

## 4) One analysis {.smaller}
- Null hypothesis: slope is equal to zero (no sig. change in lac value over time)
- Alt hypothesis: slope NOT equal to zero (lac value changes over time)
```{r, collapse=TRUE}
Lac_mod <- lm(Lac ~ Time, data =  complete_sliders)
summary(Lac_mod)
```
- Fail to reject the null hypothesis because of the high p-value for Time
- Lactate does not change significantly over time
- Time estimate value also supports the conclusion that the slope is zero

## 5) Where I am going...
- t-test comparing glass and plastic syringes?  

```{r, echo=FALSE, fig.height=7, fig.width=8}
rays_gath$Syringe <- gsub("G", "Glass", rays_gath$Syringe)
rays_gath$Syringe <- gsub("P", "Plastic", rays_gath$Syringe)

ggplot(data = rays_gath, aes(x=Time, y=Value, color=factor(Ray), by=Ray)) +
  geom_line() +
  geom_point() +
  facet_grid(Variable ~ Syringe, scales = "free")
```


## Confused.
```{r, echo=FALSE}
ggplot(data = complete_sliders, aes(x=Time, y=PCO2, color=factor(Turtle), by=Turtle)) +
  geom_line() +
  geom_point()

PCO2_mod <- lm(PCO2 ~ Time, data =  complete_sliders)
```

## Model summary
```{r, echo=FALSE, tidy=TRUE}
summary(PCO2_mod)

```

## Assumptions
```{r, fig.height=5, fig.width=5, echo=FALSE}
plot(PCO2_mod, which = 1)
plot(PCO2_mod, which = 2)
```



```{r, warning=FALSE, message=FALSE, echo=FALSE}
ggplot(data = sliders, aes(x=Time, y=Lac, color=factor(Turtle), by=Turtle)) +
  geom_line() +
  geom_point() +
  xlab("Time (minutes)") +
  ylab("Lac (mmol*L^-1") +
  ggtitle("Turtles: Lac vs. Time")

# Linear regression
Lac_mod <- lm(Lac ~ Time, data =  sliders_lac)
summary(Lac_mod)



plot(Lac_mod, which = 1)
plot(Lac_mod, which = 2)
```




```{r, include=FALSE}
ggplot(data = complete_sliders, aes(x=Time, y=Lac, color=factor(Turtle), by=Turtle)) +
  geom_line() +
  geom_point()

ggplot(data = rays_gath, aes(x=Time, y=Value, color=factor(Ray), by=Ray)) +
  geom_line() +
  geom_point() +
  facet_grid(Variable ~ Syringe, scales = "free")

ggplot(data = sliders_gath, aes(x=Time, y=Value, color=factor(Turtle), by=Turtle)) +
  geom_line() +
  geom_point() +
  facet_wrap(~Variable, scales = "free")
```

