---
title: "Rays"
output: html_document
---
```{r}
library(readxl)
library(tidyr)
library(dplyr)
library(ggplot2)


#Load
rays <- read_excel('~/Dropbox/Blood_Gas_Project/Data/BG_data_RAYS.xlsx', skip = 0)

#Work with one chunk at a time
rays_1 <- rays[1:10,1:14]

#separate meta data and labels (all columns must be named first)
  # the label "Time" will make sense after transposing data
names(rays_1) <- c("Ray", "Time", "T0", "T0", "T1", "T1", "T2", "T2", "T3", "T3", "T4", "T4", "T5", "T5")

rays_1 <- separate(data = rays_1, col = "Ray", into = c("Meta", "Data"), sep = ": ")

#save meta data in its own data frame
meta_data_rays <- data.frame(rays_1[1:4,1:2])

#take meta data out of main data frame
rays_1 <- rays_1[1:10,3:15]

#put col names in first row so they become a new column after transposing instead of just row names
rays_1[1,] <- names(rays_1)

#transpose
rays_1 <- t(rays_1)
rays_1 <- as.data.frame(rays_1, stringsAsFactors = FALSE)

#use row 1 to rename columns and clear row names
names(rays_1) <-rays_1[1,]
rays_1 <- rays_1[-1,]
row.names(rays_1) <- NULL

#transpose meta data
meta_data_rays <- t(meta_data_rays)
meta_data_rays <- as.data.frame(meta_data_rays, stringsAsFactors = FALSE)

# correct column names, clear row name in meta data
names(meta_data_rays) <- meta_data_rays[1,1:4]
meta_data_rays <- meta_data_rays[-1,]
row.names(meta_data_rays) <- NULL

# pit number, weight and wingspan in meta data as numeric
meta_data_rays$Pit_Num <- as.numeric(meta_data_rays$Pit_Num)

meta_data_rays$Weight <- as.numeric(gsub("\\skg", "", meta_data_rays$Weight))

meta_data_rays$Wingspan <- as.numeric(gsub("\\scm", "", meta_data_rays$Wingspan))

#put meta data back in main data frame
complete_rays <- data.frame(rays_1, meta_data_rays)
```


FUNCTION TO TIDY ALL RAY DATA
```{r}

read_rays_transform <- function(ray_number){
    
  #file name
  filename <- paste0('~/Dropbox/Blood_Gas_Project/Data/BG_data_RAYS.xlsx')
  
  #read in each ray
  rays <- read_excel(filename, skip = ray_number*11-11)
  
  #one ray at a time
  rays_1 <- rays[1:10,1:14]
  
  #separate meta data and labels (all columns must be named first)
  names(rays_1) <- c("Ray", "Time", "T0", "T0", "T1", "T1", "T2", "T2", "T3", "T3", "T4", "T4", "T5", "T5")
  
  rays_1 <- separate(data = rays_1, col = "Ray", into = c("Meta", "Data"), sep = ": ")
  
  #save meta data in its own data frame
  meta_data_rays <- data.frame(rays_1[1:4,1:2])
  
  #take meta data out of main data frame
  rays_1 <- rays_1[1:10,3:15]
  
  #put col names in first row so they become a new column after transposing instead of just row names
  rays_1[1,] <- names(rays_1)
  
  #transpose
  rays_1 <- t(rays_1)
  rays_1 <- as.data.frame(rays_1, stringsAsFactors = FALSE)
  
  #use row 1 to rename columns and clear row names
  names(rays_1) <-rays_1[1,]
  rays_1 <- rays_1[-1,]
  row.names(rays_1) <- NULL
  
  #transpose meta data
  meta_data_rays <- t(meta_data_rays)
  meta_data_rays <- as.data.frame(meta_data_rays, stringsAsFactors = FALSE)
  
  # correct column names, clear row name in meta data
  names(meta_data_rays) <- meta_data_rays[1,1:4]
  meta_data_rays <- meta_data_rays[-1,]
  row.names(meta_data_rays) <- NULL
  
  # pit number, weight and wingspan in meta data as numeric
  meta_data_rays$Pit_Num <- as.numeric(meta_data_rays$Pit_Num)

  meta_data_rays$Weight <- as.numeric(gsub("\\skg", "", meta_data_rays$Weight))

  meta_data_rays$Wingspan <- as.numeric(gsub("\\scm", "", meta_data_rays$Wingspan))

  #put meta data back in main data frame
  complete_rays2 <- data.frame(rays_1, meta_data_rays)
  
  #add to complete data frame
  complete_rays <- rbind(complete_rays, complete_rays2)
  return(complete_rays)
}


# Run above function for rays 2 through 10
for(i in 2:10){
  complete_rays <- read_rays_transform(i)
}

# cut out non-focal variables
complete_rays <- complete_rays[,-c(3:6)]

# Filter out rows with NAs for all 4 variables (Lac, PO2, PCO2, and pH)
complete_rays <- complete_rays[!with(complete_rays,is.na(Lac)& is.na(PO2)& is.na(PCO2)& is.na(pH)),]

# Convert time to numeric values (in minutes) to plot accurately (since the T0-T5 integers are not constant intervals).
complete_rays$Time <- gsub("T0", 0, complete_rays$Time)
complete_rays$Time <- gsub("T1", 5, complete_rays$Time)
complete_rays$Time <- gsub("T2", 10, complete_rays$Time)
complete_rays$Time <- gsub("T3", 15, complete_rays$Time)
complete_rays$Time <- gsub("T4", 45, complete_rays$Time)
complete_rays$Time <- gsub("T5", 90, complete_rays$Time)

# need to numeric for temp corrections
complete_rays$pH <- as.numeric(complete_rays$pH)
complete_rays$PCO2 <- as.numeric(complete_rays$PCO2)
complete_rays$PO2 <- as.numeric(complete_rays$PO2)

complete_rays$Lac <- as.numeric(complete_rays$Lac)

# temp corrections
complete_rays$pH <- (0.795*complete_rays$pH) + 1.797
complete_rays$PCO2 <- (0.173*complete_rays$PCO2) + 0.775
complete_rays$PO2 <- (0.572*complete_rays$PO2) - 1.449



#write.csv(complete_rays, file = 'focal_ray_data.csv')

