---
title: "super turtle function"
output: html_document
---
The objective is to see how long after taking a blood sample can the blood acid-base parameters be measured accurately or if there is

1) monitor the changes in blood acid-base parameters as a function of time
2) observe the difference in these values between different syringe types.


DATA TIDYING FOR FIRST TURTLE DATA 
```{r}
library(readxl)
library(tidyr)
library(dplyr)
library(ggplot2)
library(censReg)


#Load
sliders <- read_excel('~/Dropbox/Blood_Gas_Project/Data/BG_data_SLIDERS.xlsx', skip = 0)

#work with only one chunk at a time
sliders_1 <-sliders[1:9, 1:8]

#separate meta data labels and values
names(sliders_1) <- c("Turtle", "Value", "T0", "T1", "T2", "T3", "T4", "T5")

sliders_1 <- separate(data = sliders_1, col = "Turtle", into = c("Meta", "Data"), sep = ": ")

#save it in its own data frame and remove from table for now
meta_data_sliders <- data.frame(sliders_1[1:4, 1:2])

sliders_1 <- sliders_1[1:9,3:9]
sliders_1 <- sliders_1[-1,]

# TRANSPOSE
sliders_1 <- t(sliders_1)
sliders_1 <- as.data.frame(sliders_1, stringsAsFactors = FALSE)

#add time column, remove row names, and fix the column names
sliders_1$Time <- gsub("Value", "Time", rownames(sliders_1))
rownames(sliders_1) <- NULL
names(sliders_1) <- sliders_1[1,1:9]
sliders_1 <- sliders_1[-1,]

# TRANSPOSE AND FORMAT META DATA
meta_data_sliders <- t(meta_data_sliders)
meta_data_sliders <- as.data.frame(meta_data_sliders, stringsAsFactors = FALSE)

# correct column names, clear row name in meta data
names(meta_data_sliders) <- meta_data_sliders[1,1:4]
meta_data_sliders <- meta_data_sliders[-1,]
rownames(meta_data_sliders) <- NULL

# turtle, temp, and weight values as numeric
meta_data_sliders$`Turtle` <- as.numeric(meta_data_sliders$`Turtle`)

meta_data_sliders$`Internal Temp` <- as.numeric(gsub("\\sC", "",
                                          meta_data_sliders$`Internal Temp`))
meta_data_sliders$`Weight` <- as.numeric(gsub("\\sg", "",
                                              meta_data_sliders$`Weight`))

#put meta data back onto main data frame
complete_sliders <- data.frame(sliders_1, meta_data_sliders)
``` 

FUNCTION TO TRANSFORM/TIDY AND RBIND THE REST OF THE TURTLES
```{r}
library(readxl)
library(tidyr)
library(dplyr)
library(ggplot2)

# FUNCTION:

read_sliders_transform <- function(turtle_number){
  #file name
  filename <- paste0('~/Dropbox/Blood_Gas_Project/Data/BG_data_SLIDERS.xlsx')
  
  #read in per turtle
  sliders <- read_excel(filename, skip = turtle_number*10-10)
  
  #only one turtle at a time
  sliders_1 <-sliders[1:9, 1:8]
  
  #rename columns
  names(sliders_1) <- c("Turtle", "Value", "T0", "T1", "T2", "T3", "T4", "T5")
  
  #separate meta data into two columns
  sliders_1 <- separate(data = sliders_1, col = "Turtle", into = c("Meta", "Data"), sep = ": ")
  
  #save meta data as a new data frame
  meta_data_sliders <- data.frame(sliders_1[1:4, 1:2])
  
  #get rid of extra rows and columns
  sliders_1 <- sliders_1[2:9,3:9]
  
  #transpose
  sliders_1 <- t(sliders_1)
  
  #get rid of factors
  sliders_1 <- as.data.frame(sliders_1, stringsAsFactors = FALSE)
  
  #add time column and rename value
  sliders_1$Time <- gsub("Value", "Time", rownames(sliders_1))
  
  #rename columns
  names(sliders_1) <- sliders_1[1,1:9]
  
  #remove first row
  sliders_1 <- sliders_1[-1,]
  
  # TRANSPOSE AND FORMAT META DATA
  meta_data_sliders <- t(meta_data_sliders)
  meta_data_sliders <- as.data.frame(meta_data_sliders, stringsAsFactors = FALSE)
  
  # correct column names
  names(meta_data_sliders) <- meta_data_sliders[1,1:4]
  meta_data_sliders <- meta_data_sliders[-1,]
  
  #clear row names
  rownames(meta_data_sliders) <- NULL
  
  # Make variables numeric and remove extra characters
  meta_data_sliders$`Turtle` <- as.numeric(meta_data_sliders$`Turtle`)
  meta_data_sliders$`Internal Temp` <- as.numeric(gsub("\\sC", "", meta_data_sliders$`Internal Temp`))
  meta_data_sliders$`Weight` <- as.numeric(gsub("\\sg", "", meta_data_sliders$`Weight`))
  
  #add meta data back to data frame
  complete_sliders2 <- data.frame(sliders_1, meta_data_sliders)
  
  #rbind to compelte data frame
  complete_sliders <- rbind(complete_sliders, complete_sliders2)
  return(complete_sliders)
}
  
# Run above function for turtles 2 through 10
for(i in 2:10){
  complete_sliders <- read_sliders_transform(i)
}


#need to be numeric for temperature corrections
complete_sliders$pH <- as.numeric(complete_sliders$pH)
complete_sliders$PCO2 <- as.numeric(complete_sliders$PCO2)
complete_sliders$PO2 <- as.numeric(complete_sliders$PO2)

#temperature corrections
complete_sliders$temp_change <- (37 - complete_sliders$Internal.Temp)
complete_sliders$pH <- ((0.0144*complete_sliders$temp_change) + complete_sliders$pH)
complete_sliders$PCO2 <- (complete_sliders$PCO2*(10^-0.019*complete_sliders$temp_change))
complete_sliders$PO2 <- (complete_sliders$PO2*(10^-0.0058*complete_sliders$temp_change))

# Filter NAs in pH 
  #(no values for turtle 3 at T4, removes whole row of NAs)
complete_sliders <- complete_sliders %>%
  filter(!is.na(pH))

# time to numeric values (in minutes)
complete_sliders$Time <- gsub("T0", 0, complete_sliders$Time)
complete_sliders$Time <- gsub("T1", 5, complete_sliders$Time)
complete_sliders$Time <- gsub("T2", 10, complete_sliders$Time)
complete_sliders$Time <- gsub("T3", 15, complete_sliders$Time)
complete_sliders$Time <- gsub("T4", 45, complete_sliders$Time)
complete_sliders$Time <- gsub("T5", 90, complete_sliders$Time)

# make time numeric
complete_sliders$Time <- as.numeric(complete_sliders$Time)


```

DATA FRAME WITH ONLY FOCAL VARIABLES AND WITHOUT META DATA
```{r, eval=FALSE}
#store all variables and meta data before cutting out extra
tidy_complete_sliders <- complete_sliders

#cut out non-focal variables for analysis
# (keep Lac, pH, PCO2, PO2)
complete_sliders <- complete_sliders[1:59,5:10]


```


***
ANALYSIS

```{r}
##### Lac #####


# assume <0.30 = 0 for Lac values for Turtle 10 for analysis
complete_sliders$Lac <- as.numeric(gsub("<0.30", "0", complete_sliders$Lac))


ggplot(data = complete_sliders, aes(x = Time, y = Lac)) +
  geom_point() +
  scale_y_continuous() +
  stat_smooth(method = lm)
  

complete_sliders$Lac2 <- as.numeric(complete_sliders$Lac)

Lac_mod <- lm(Lac2 ~ Time, data =  complete_sliders)
summary(Lac_mod)

complete_sliders$Lac <- as.numeric(complete_sliders$Lac)



##### pH #####
ggplot(data = complete_sliders, aes(x = Time, y = pH)) +
  geom_point() +
  stat_smooth(method = lm)

pH_mod <- lm(pH ~ Time, data = complete_sliders)
summary(pH_mod)




##### PCO2 #####

ggplot(data = complete_sliders, aes(x = Time, y = PCO2, color = factor(Turtle), by = Turtle)) +
  geom_line() +
  geom_point() +
  stat_smooth(method = lm)

complete_sliders$PCO2_log <- log(complete_sliders$PCO2)

PCO2_mod <- lm(sqrt(PCO2) ~ Time, data = complete_sliders %>% filter(Time<25))
#####VARIABILITY MENTION#####
summary(PCO2_mod)
plot(PCO2_mod)

##### PO2 #####
library(car)
install.packages('pbkrtest')

# Plot PO2 by turtle over time
ggplot(data = complete_sliders, aes(x = Time, y = PO2, color = factor(Turtle), by = Turtle)) +
  geom_line() +
  geom_point() +
  stat_smooth(method = lm)

  # Log transform PO2
complete_sliders$log_PO2 <- log10(complete_sliders$PO2)

  # Plot log transformed PO2 vs Time data
ggplot(data = complete_sliders, aes(x = Time, y = log_PO2, color = factor(Turtle), by = Turtle)) +
  geom_line() +
  geom_point()

  # Model
PO2_mod <- lm(log(PO2) ~ Time, data = complete_sliders)
summary(PO2_mod)
plot(PO2_mod)

PO2_mod2 <- lm(PO2 ~ Time, data = complete_sliders)
summary(PO2_mod2)
plot(PO2_mod2)



```


```{r, eval=FALSE}

censReg(formula = Lac ~ Time, data = complete_sliders)


## Kleiber & Zeileis ( 2008 ), page 142
data( "Affairs", package = "AER" )
estResult <- censReg( affairs ~ age + yearsmarried + religiousness +
   occupation + rating, data = Affairs )
print( estResult )


## Kleiber & Zeileis ( 2008 ), page 143
estResultBoth <- censReg( affairs ~ age + yearsmarried + religiousness +
   occupation + rating, data = Affairs, right = 4 )
print( estResultBoth )


# gamma distribution
# family = gamma (link = log)
```

```{r, eval=FALSE}

# write.csv(tidy_sliders, file = 'tidy_turtle_data.csv')
# write.csv(complete_sliders, file = 'focal_turtle_data_zeros.csv')
```



TURTLE 6 T3 = 18 MIN NOT 15


